---
title: "Adding Your Data"
author: "Joy Payton"
date: "2/26/2019"
output: 
  html_document: 
    toc: yes
    toc_depth: 4
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE, cache = TRUE)
```

## Purposes

So, you have some data... perhaps you want to look at how your patients are responding to an exercise intervention for obesity and see if there's a difference in compliance and weight loss success that could be related to violence near home that make it unsafe to play outside.  

## Obtain Geographic Data

The City of Philadelphia supplies information about shootings (including officer-involved shootings) which includes data about the shooting victim and the location.  Here, we're really interested in the location of shootings over the past few years, to understand what parts of Philadelphia are more prone to this specific kind of violence.

To see more information about this dataset, please visit <https://www.opendataphilly.org/dataset/shooting-victims/resource/a6240077-cbc7-46fb-b554-39417be606ee?inner_span=True>.

For our purposes, we're going to get the bare minimum of information: latitude and longitude.  The API endpoint is described in the link above and uses a SQL query to select only the data we care about.  Because our query has spaces and other special characters, we need to "encode" it for request.

The data will come in as json, which we'll parse.

```{r shooting-data}
library(jsonlite)
url <- URLencode('https://www.opendataphilly.org/api/action/datastore_search_sql?sql=SELECT lat, lng from "a6240077-cbc7-46fb-b554-39417be606ee"')
shooting_data <- fromJSON(url)
```

We can examine the shooting data by using R's `str` (structure) command:

```{r}
str(shooting_data)
```

Here we see that we have a data frame, accessible at `shooting_data$result$records`:

```{r}
head(shooting_data$result$records, 6)
```

## Mapping Points

If we wanted to, we could easily create a map of these shootings, just based on latitude and longitude.  Since latitude and longitude are currently in "chr" (character) format, we'll make them numeric so that we can do math on them.  We'll create a map that's centered on the mean latitude and longitude of all our shootings, and which is appropriately zoomed in (you might have to experiment with the zoom factor).

We're going to add a standard road map below to show more context, using `addTiles`.

```{r}
library(leaflet)
library(leaflet.extras)
library(dplyr)

shootings <- shooting_data$result$records
shootings$lat <- as.numeric(shootings$lat)
shootings$lng <- as.numeric(shootings$lng)

shootings %>% 
  leaflet() %>% 
  addTiles() %>%
  setView(lng = mean(as.numeric(shootings$lng), na.rm=TRUE), 
          lat = mean(as.numeric(shootings$lat), na.rm=TRUE), zoom = 10) %>%
  addMarkers(clusterOptions = markerClusterOptions()) %>%
  suspendScroll()
```

## Mapping Polygons

What's more likely, however, is that we want to use polygon data to create a map that shows how much a particular area is affected.  This is because we want to create *combined* data -- we want to put information about our patients or research subjects along with the level of violence they are exposed to.  Instead of using latitude and longitude, we'll gather the number of shootings per Census tract, which we can then use as a proxy for violence exposure for the patients and subjects who live in that Census tract.  It's a sort of "binning", but using the existing "bins" of Census tracts.  

Let's start by getting a map:

```{r}
library(rgdal)
philadelphiaCensusTracts <- readOGR("http://data.phl.opendata.arcgis.com/datasets/8bc0786524a4486bb3cf0f9862ad0fbf_0.geojson")
```

#### Mapping Point Data to Polygons

Now what we'd like to do is get the shootings-per-tract data, which we can then combine with our research or clinical data to see if violence near home has any effect on our outcomes.  To do this, we take the latitude and longitude of our shootings and transform them slightly so that they are understood as spatial coordinates, not just pairs of numbers.  We'll use the same map projection used in our original `philadelphiaCensusTracts`.

```{r}
library(sp)
coordinates <- SpatialPoints(shootings[c("lng", "lat")])
proj4string(coordinates) <- proj4string(philadelphiaCensusTracts)
```

Let's now apply what we know about our polygons (from `philadelphiaCensusTracts`) and apply that to our points.  We'll end up with a table that has one row for each shooting coordinate.  Essentially, what we're doing is taking each point, lining it up with a matching polygon, and then getting the data about that polygon, which came along with the geoJSON file we downloaded.

```{r}
shooting_tract_data <- over(coordinates, philadelphiaCensusTracts)
head(shooting_tract_data)
```

We see the first few lines of the Census data for each of our shootings.  For example, the first shooting in our shooting data corresponds to Census tract 36, which is in State 42 (Pennsylvania) and County 101 (Philadelphia County).  We can use this to find out how many shootings take place in each Census tract.  


```{r}
shootings_by_census_tract <- shooting_tract_data %>% 
                                 group_by(GEOID10) %>% 
                                 summarise(num_shootings = n()) %>% 
                                 ungroup() 
head(shootings_by_census_tract)
```

#### Handling Empty Data

Don't forget that there are some Census tracts that aren't represented at all in our `shooting_tract_data` data frame, so let's make sure we enrich it with all the tracts that aren't included in the shooting data.  We can get those by taking the data frame of our tract data, selecting the list of all the Census tracts in Philadelphia, and making sure that if they weren't mentioned above, we add them, but with `num_shootings` equal to 0.

```{r}
non_shooting_tracts <- philadelphiaCensusTracts@data %>% 
                       select(GEOID10) %>%
                       filter(!GEOID10 %in% shootings_by_census_tract$GEOID10) %>%
                       mutate(num_shootings = 0)
head(non_shooting_tracts)
```

We can now combine the tracts-with-shootings and the tracts-with-no-shootings to get an overall picture of violence by census tract:

```{r}
shootings_by_census_tract <- rbind(shootings_by_census_tract, non_shooting_tracts)
```

Let's map this!  We need to combine the data we aggregated with the data in the map file.  

***WARNING: PITFALL AHEAD!***

We have a really simple merge, right?  Census tract information is contained in both datasets, so we have a "hinge":

```{r}
merged_data <- merge (philadelphiaCensusTracts@data, shootings_by_census_tract, by = "GEOID10")
```

Let's peek!

```{r}
head(merged_data) 
```
Looks great, and as `merge` is wont to do, it's ordered the new data frame by the merge variable, `GEOID10`.

Can we just put that back into the map?

No!  While maintaining the order of rows is not something we're used to doing in R, we absolutely have to maintain the original order of rows when we're using spatial polygon data.  That's because the rows of the data frame correspond 1:1, **in order**, to the polygons.

There are a few ways to work around this problem, but we have a great, easy to use solution.  Let's take a peek at our original order:

```{r}
head(philadelphiaCensusTracts@data)
```

Yes, there's a great "OBJECTID" column that's numbered incrementally.  So it's very simple for us to take our merged dataframe and re-instate that order:

```{r}
merged_data <- merged_data[order(merged_data$OBJECTID),]
head(merged_data)
```

Now, **this** data frame we can add back into the spatial polygon data frame safely:

```{r}
philadelphiaCensusTracts@data <- merged_data
```

And then mapping is a snap:

```{r}
shooting_palette <- colorBin("Reds", domain = philadelphiaCensusTracts@data$num_shootings, bins = 5, na.color = "#808080")

leaflet() %>%
    addPolygons(
    data = philadelphiaCensusTracts,
    fillColor = ~shooting_palette(philadelphiaCensusTracts@data$num_shootings),
    weight = 1,  # border thickness
    opacity = 0.5, # border opacity
    color = "white", # border color
    fillOpacity = 1,
    label = paste("Number of Shootings: ", philadelphiaCensusTracts@data$num_shootings, sep = "")) %>%
  suspendScroll()
```

We could also plot the actual shootings on top -- that might be duplicative effort but it's also helpful to make sure there's nothing obviously wrong in the mapping.

```{r}
leaflet() %>%
    addPolygons(
    data = philadelphiaCensusTracts,
    fillColor = ~shooting_palette(philadelphiaCensusTracts@data$num_shootings),
    weight = 1,  # border thickness
    opacity = 0.5, # border opacity
    color = "white", # border color
    fillOpacity = 1,
    label = paste("Number of Shootings: ", philadelphiaCensusTracts@data$num_shootings, sep = "")) %>%
  addCircles(lat = shootings$lat, lng = shootings$lng, color = "dark grey", weight = 1, radius = 100) %>%
  suspendScroll()
```

So what would that map look like if we hadn't handled the row ordering problem?  Let's compare:

```{r}
badPhilly <- philadelphiaCensusTracts
badPhilly@data <- merged_data[order(merged_data$GEOID10),]


leaflet() %>%
    addPolygons(
    data = badPhilly,
    fillColor = ~shooting_palette(badPhilly@data$num_shootings),
    weight = 1,  # border thickness
    opacity = 0.5, # border opacity
    color = "white", # border color
    fillOpacity = 1,
    label = paste("Number of Shootings: ", 
                  badPhilly@data$num_shootings, sep = "")) %>%
  addCircles(lat = shootings$lat, lng = shootings$lng, color = "dark grey", weight = 1, radius = 100) %>%
  suspendScroll()
```

You can see that the extra diagnostic layer was helpful to show that the shooting data points don't line up with the polygons.

## Adding Some Proprietary Data To The Mix

***Important aside: this data is a work of fiction, so: ***

> Any resemblance to actual persons, living or dead, or actual events is purely coincidental.

Let's take a peek at our fake data, which purports to give the number of minutes various patients on an obesity-related research project reported exercising.

```{r}
fake_exercise_data <- read.csv("../Data/fake_exercise_data.csv", stringsAsFactors = FALSE)
head(fake_exercise_data)
```

```{r}
exercise_per_tract <- fake_exercise_data %>% 
                      group_by(census_tract) %>%
                      summarise(mean_exercise = mean(daily_exercise_minutes)) %>%
                      ungroup()
head(exercise_per_tract)
```


### Combine with Map Data Frame!

***WARNING WARNING PITFALL AHEAD!***

First, we'll combine `exercise_per_tract` with the data found in `philadelphiaCensusTracts@data`.  This time, we're combining the short name, not the fully qualified GEOID (who knows why?!).

```{r}
census_tracts <- merge(x=philadelphiaCensusTracts@data, y=exercise_per_tract, by.x="NAME10", by.y="census_tract", all.x = TRUE)
```

Then we'll add our enriched data back to the geojson data, so that in addition to the fields it came with, it will now contain the exercise and shooting data we gathered.  It's important to order this data by the OBJECTID so that the correct polygon is associated with the correct data!

```{r}
philadelphiaCensusTracts@data <- census_tracts[order(census_tracts$OBJECTID),]
```

Now, let's create an interactive map!  We'll color the polygons by exercise amount to begin with.

```{r fig.width = 10}
exercise_palette <- colorBin("Blues", domain = philadelphiaCensusTracts@data$mean_exercise, bins = 8, na.color = "#cccccc")

interactive_map <- leaflet(philadelphiaCensusTracts) %>%
  addPolygons(
    data = philadelphiaCensusTracts,
    fillColor = ~exercise_palette(philadelphiaCensusTracts@data$mean_exercise),
    weight = 1,  # border thickness
    opacity = 0.5, # border opacity
    color = "white", # border color
    fillOpacity = 1) %>%
  suspendScroll()
interactive_map 
```

Now, let's add some labels.  We'll do variable interpolation to create labels that tell what each Census tract is and the exercise and shooting data for that tract:

```{r}
labels <- sprintf(
  "<strong>%s</strong><br/>
  Exercise in Minutes: %g <br/>
  Number of Shootings: %g",
  philadelphiaCensusTracts@data$NAMELSAD10, 
  philadelphiaCensusTracts@data$mean_exercise,
  philadelphiaCensusTracts@data$num_shootings
) %>% lapply(htmltools::HTML)
```



Then we'll create the map again, but with labels.  This allows the viewer to see at a glance the violence and exercise metrics for each tract!

```{r}
interactive_map <- leaflet(philadelphiaCensusTracts) %>%
  setView(lng = mean(as.numeric(shootings$lng), na.rm=TRUE), 
          lat = mean(as.numeric(shootings$lat), na.rm=TRUE), zoom = 11) %>%
  addPolygons(
    data = philadelphiaCensusTracts,
    fillColor = ~exercise_palette(philadelphiaCensusTracts@data$mean_exercise),
    weight = 1,  # border thickness
    opacity = 0.5, # border opacity
    color = "white", # border color
    fillOpacity = 1,
    label = labels,
  labelOptions = labelOptions(
    style = list("font-weight" = "normal", padding = "3px 8px"),
    textsize = "15px",
    direction = "auto")) %>%
  suspendScroll()
interactive_map 
```

## Combine it All!

So, we have a couple of polygon maps and a point map -- how can we combine them all in Leaflet?  We're going to use a layer controller.

```{r}
my_map <- leaflet() %>%
  setView(lng = mean(philadelphiaCensusTracts@bbox['x',], na.rm=TRUE), 
          lat = mean(philadelphiaCensusTracts@bbox['y',], na.rm=TRUE), zoom = 11) %>%
addPolygons(data = philadelphiaCensusTracts,
    weight = 1,  # border thickness
    opacity = 0.5, # border opacity
    color = "grey", # border color
    fillColor = ~shooting_palette(philadelphiaCensusTracts@data$num_shootings),
    fillOpacity = 1,
    label = labels,
    labelOptions = labelOptions(
    style = list("font-weight" = "normal", padding = "3px 8px"),
    textsize = "15px",
    direction = "auto"),
    group = "Frequency of Shootings With Injury"
  ) %>%
    addPolygons(
    data = philadelphiaCensusTracts,
    weight = 1,  # border thickness
    opacity = 0.5, # border opacity
    color = "grey", # border color
    fillColor = ~exercise_palette(philadelphiaCensusTracts@data$mean_exercise),
    fillOpacity = 1,
    label = labels,
    labelOptions = labelOptions(
    style = list("font-weight" = "normal", padding = "3px 8px"),
    textsize = "15px",
    direction = "auto"),
      group = "Mean Exercise in Minutes"
  ) %>%
  addCircles(lat = shootings$lat, lng = shootings$lng, color = "dark grey", weight = 1, radius = 100, group = "Frequency of Shootings With Injury") %>%
    addCircles(lat = shootings$lat, lng = shootings$lng, color = "dark grey", weight = 1, radius = 100, group = "Mean Exercise in Minutes") %>%
    addLayersControl(
    baseGroups = c("Frequency of Shootings With Injury", "Mean Exercise in Minutes"),
    options = layersControlOptions(collapsed = FALSE)
  ) %>%
  suspendScroll()

my_map
```
```

