---
title: "CensusAPI"
author: "Joy Payton"
date: "3/13/2019"
output: 
  html_document: 
    toc: yes
    toc_depth: 4
    toc_float: yes
---

```{r setup, include=FALSE}

# Note you may need to install some packages!
# install.packages(c("kableExtra", "tidycensus"), dependencies = TRUE)

# You will CERTAINLY need to get a census API key -- read on below.

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, cache = TRUE)
library(kableExtra)
```

## APU Setup

First, I'm going to pull in my API key.  You'll want to change the line below or put a file containing your API key in the place specified below.  I've excluded the "Private" folder from being included in the GitHub repo, so you'll want to add your own on your local machine.

```{r}
census_key <- readLines("../Private/census_api_key.txt")
```

Let's use `tidycensus`!  It creates a smoother experience for a handful of Census API calls.  I'll set my census key once using `tidycensus::census_key` so that I don't have to keep referencing it in API calls I make.

```{r}
library(tidycensus)
census_api_key(census_key)
```

## American Community Survey

What variables are available, say, for the five year American Community Survey?  I need to know the end year of the survey I care about, so I can go to <https://www.census.gov/programs-surveys/acs> to discover that 2017 is the latest year end.  Scroll through the variables below ... there are a LOT.

```{r}
acs5_vars <- load_variables(2017, "acs5")
kable(acs5_vars) %>%
  kable_styling() %>%
  scroll_box(width = "100%", height = "300px")
```

There are `r nrow(acs5_vars)` variables that describe `r length(unique(acs5_vars$concept))` unique concepts.

Let's take a look at the concepts that are available:

```{r}
kable(unique(acs5_vars$concept)) %>%
  kable_styling() %>%
  scroll_box(width = "100%", height = "300px")
```

Maybe we want to hone in on, say, median household income for the census tracts that comprise New York City.  Specifically, I want the variable `B19113_001`.  In my use of `get_acs` I can use geographical names, like "New York", or the FIPS code for those places.  Here, to foster ease of reading, I've used the county names of the five boroughs.

```{r}
nyc_median_income <- get_acs(geography = "tract",
                             variables = "B19113_001",
                             state = "NY",
                             county = c("New York", "Richmond", "Queens", "Kings", "Bronx"),
                             survey = "acs5")
```

There are `r nrow(nyc_median_income)` census tracts represented!  Let's take a peek at this data frame and get some summary statistics.

```{r}
kable(head(nyc_median_income)) %>%
  kable_styling()
kable(summary(nyc_median_income)) %>%
  kable_styling()
```

Why are some of my rows missing data?  They could be tracts that correspond to areas like parks or rivers where there's only a transient population and there's no meaningful way to make predictions about median income.  When we map this data, we might understand more.

